---
title: "Lab 03 - Using Propensity Scores"
output: github_document
author: "Ellie Grace"
date: '2022-03-29'
---


```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(smd)
library(cli)
library(gtsummary)
library(ggdag)
library(survey)
library(dagitty)
library(broom)
```

## Exercise 1

```{r}
coffee <- read.csv("coffee_town_df.csv")

coffee %>% count(coffee_town)
```
For this data set we have 1100 observations with coffee_town as our exposure. There are 987 in the unexposed group (those who do not live in coffee town) and 113 in the exposed group (those who do live in coffee town). The outcome is the cups_of_coffee and the other variables are job, smoke, and age. Job is categorical and ranks the jobs level of difficulty as either easy, hard, or none. Smoke is also a categorical variable with categories never, former, or current. There is no missing data.

## Exercise 2

```{r}
coffee %>%
  tbl_summary(
    by = coffee_town, 
    include = c( #Vector of adjustment variables
                "cups_of_coffee",
                "age",
                "job",
                "smoke")) %>%
  add_overall()
```

From this table, we are able to get a better idea of how the data is distributed. For example, it is more common for those who do not live in coffee town to drink 1-3 cups of coffee than those who do live in coffee town. We can also notice that 0% of people who do not live in coffee town drink 5-6 cups of coffee whereas 3.5% and .9% of people who live in coffee town drink 5 and 6 cups of coffee. We can also notice that coffee town has a higher median age and it looks like the ages have a higher range (although skewed on the older side) than the ages of those who do not live in coffee town. Looking at the job section, we can see that coffee town has a higher unemployment rate whereas those who do not live in coffee town tend to have easier jobs. Lastly, the smoking categories between the exposure groups appear to be relatively similar. Majority of people in both places have never smoked, although there are more current smokers not living in coffee town. 

## Exercise 3

```{r}
coffee <- 
  glm(coffee_town ~ age + smoke + job,
    data = coffee,
    family = binomial()) %>%
  augment(type.predict = "response", data = coffee) 

df_plot_ps <- coffee %>%
  tidyr::pivot_wider(names_from = coffee_town, 
                     values_from = .fitted, 
                     names_prefix = "coffee_town_p")

ggplot(df_plot_ps) +
  geom_histogram(bins = 50, 
                 aes(x = coffee_town_p1), 
                 fill = "deeppink2") + 
  geom_histogram(bins = 50, 
                 aes(x = coffee_town_p0, 
                     y = - stat(count)), 
                 fill = "plum4") + 
  
  scale_y_continuous("count", label = abs) + 
  scale_x_continuous("propensity score") +
  annotate(geom = "label",
           x = 0.6,
           y= 20,
           label = "Lives in Coffee Town") + 
annotate(geom = "label",
         x = 0.5,
         y = -40,
         label  = "Doesn't Live in Coffee Town") +
  ggtitle("Distribution of Propensity Scores")
  theme_minimal()
```

We see that our model did a good job for those who do not live in coffee town (since they all have low probabilities of living in coffee town), but it is a little more scattered for coffee town. We see for the coffee town side of the histogram there are some with low probabilities of living there, some with probabilities around .3, and few with high probabilities of living there. This could perhaps be due to the small number of observations in the exposed group.

## Exercise 4

```{r}
coffee <- coffee %>%
  mutate( 
    w_ato = (1-.fitted) * coffee_town + .fitted * (1-coffee_town)
  )

df_plot_ps <- coffee %>%
  tidyr::pivot_wider(names_from = coffee_town, 
                     values_from = .fitted, 
                     names_prefix = "coffee_town_p")
```

The causal estimand we will be estimating is the average treatment effect among the overlap. 

## Exercise 5

```{r}
svy_des <- svydesign(
  ids = ~ 1,
  data = coffee,
  weights = ~ w_ato
)

svy_des %>%
  tbl_svysummary(by = coffee_town,
              include = c("age", 
      "job",
      "smoke",
      "cups_of_coffee")) %>%
  add_overall()
```

We can first notice that we have a much smaller amount of people now compared to our unweighted table. Then, none of the columns directly match the overall, exposed, or unexposed group (which is expected since we are looking at the overlap population). Although the job and smoking sections are closest to the unweighted exposed group. Lastly, the patterns of the cups of coffee section roughly match the same patterns found in the unweighted table. 

## Exercise 6

```{r}
ggplot(df_plot_ps) +
  geom_histogram(bins = 50, 
                 alpha = 0.5,
                 aes(x = coffee_town_p1), 
                 fill = "deeppink2") + 
  geom_histogram(bins = 50, 
                 aes(x = coffee_town_p1,  # Treatment arm
                     weight = w_ato), 
                 fill = "deeppink2") +
#Control arm
  geom_histogram(bins = 50, 
                 aes(x = coffee_town_p0, 
                     y = - stat(count)),
                 fill = "plum4",
                 alpha = 0.5) + 
  geom_histogram(bins = 50,
                 aes(x = coffee_town_p0, 
                     weight = w_ato, 
                     y = - stat(count)), 
                 fill = "plum4") + 
  
  scale_y_continuous("Count", label = abs) + 
  scale_x_continuous("Propensity Score") +
  annotate(geom = "label",
           x = 0.5,
           y= 20,
           label = "Lives in Coffee Town") +
annotate(geom = "label",
         x = 0.25,
         y = -40,
         label  = "Doesn't Live in Coffee Town") + 
  ggtitle("Distribution of Weighted/Unweighted Propensity Scores") + 
  theme_minimal()
```

For this histogram, since we are using the overlap weight, an observation will get penalized more if they are in the group they are suspected to be in. But if an observation is in a "nonintuitive location"--i.e. those who have a low probability of living in coffee town but do--then they will not be penalized as much seeing as they are most likely in the overlap population. 

## Exercise 7

```{r}
smds <- coffee %>%
  summarize(
    across(c(age, job, smoke),
           list(unweighted = ~smd(.x, coffee_town)$estimate,
                ato.weighted = ~smd(.x, coffee_town, w = w_ato)$estimate))
  )

plot_df <- smds %>%
  pivot_longer(
    everything(),
    values_to = "SMD",
    names_to = c("Variable", "Method"),
    names_pattern = "(.*)_(.*)"
  ) %>%
  arrange(Method, abs(SMD)) %>%
  mutate(Variable = fct_inorder(Variable))

ggplot(
  data = plot_df,
  aes(x = abs(SMD), y = Variable, 
      group = Method, color = Method)
) +  
  geom_line(orientation = "y") +
  geom_point() + 
  geom_vline(xintercept = 0.1, 
             color = "black", size = 0.1) +
  ggtitle("Love Plot for Smoke, Age, and Job") + 
  theme_minimal()
```

For the unweighted smd's, we are able to see that both smoke and age have a higher value than we would like, and job has a much higher value than we would like. Now for the ATO weighted smd's, we are able to see that our weights balanced on the mean and now are perfectly centered at zero. 

## Exercise 8

```{r}
ggplot(coffee, 
       aes(x = age, group = coffee_town, 
           color = factor(coffee_town))) +
  stat_ecdf() +
  scale_color_manual("Lives in Coffee Town", 
                     values = c("plum4", "deeppink2"),
                     labels = c("No", "Yes")) + 
  scale_x_continuous("Age") + 
  ylab("Proportion <= x") + 
  ggtitle("Unweighted eCDF for Age") +
  theme_minimal()
```
We already knew that coffee town had a wider range of ages, but now we can see that coffee town has only young people and old people whereas the ages of those who do not live in coffee town appear more evenly distributed. 

```{r}
ecdf_1 <- coffee %>%
  filter(coffee_town == 1) %>%
  arrange(age) %>%
  mutate(cum_pct = cumsum(w_ato) / sum(w_ato))

ecdf_0 <- coffee %>%
  filter(coffee_town == 0) %>%
  arrange(age) %>%
  mutate(cum_pct = cumsum(w_ato) / sum(w_ato))

ggplot(ecdf_1, aes(x = age, y = cum_pct)) +
  geom_line(color = "deeppink2") +
  geom_line(data = ecdf_0, 
            aes(x = age, y = cum_pct), 
            color = "plum4") + 
  xlab("Age") + 
  ylab("Proportion <= x") + 
  ggtitle("Age after ATO Weighting") +
annotate(geom = "label",
         x = 30,
         y = 0.05,
         color = "plum4",
         label = "Doesn't Live in Coffee Town") +
annotate(geom = "label",
         x = 20,
         y = 0.45,
         color = "deeppink2",
         label = "Lives in Coffee Town") +
  theme_minimal()
```
After ATO weighting, the distributions match a little more closely, but still do not line up well.

## Exercise 9

We will now iterate our propensity score model and fit age with a natural spline. Hopefully doing this will result in a better looking eCDF!

```{r}
coffee_2 <- glm(coffee_town ~ job + smoke +
      splines::ns(age, 3),
    family = binomial(),
    data = coffee) %>%
  augment(data = coffee, type.predict = "response") 

coffee_2 <- coffee_2 %>%
  mutate( 
    w_ato = (1-.fitted) * coffee_town + .fitted * (1-coffee_town)
  )
```

```{r}
df_plot_ps <- coffee_2 %>%
  tidyr::pivot_wider(names_from = coffee_town, 
                     values_from = .fitted, 
                     names_prefix = "coffee_town_p")

ggplot(df_plot_ps) +
  geom_histogram(bins = 50, 
                 aes(x = coffee_town_p1), 
                 fill = "deeppink2") + 
  geom_histogram(bins = 50, 
                 alpha = 0.5, 
                 aes(x = coffee_town_p1,  # Treatment arm
                     weight = w_ato), 
                 fill = "deeppink2") +
#Control arm
  geom_histogram(bins = 50, 
                 aes(x = coffee_town_p0, 
                     y = - stat(count)),
                 fill = "plum4",
                 alpha = 0.5) + 
  geom_histogram(bins = 50,
                 aes(x = coffee_town_p0, 
                     weight = w_ato, 
                     y = - stat(count)), 
                 fill = "plum4") + 
  
  scale_y_continuous("Count", label = abs) + 
  scale_x_continuous("Propensity Score") +
  annotate(geom = "label",
           x = 0.75,
           y= 60,
           label = "Lives in Coffee Town") +
annotate(geom = "label",
         x = 0.5,
         y = -60,
         label  = "Doesn't Live in Coffee Town") + 
  ggtitle("Weighted Propensity Scores after Iterating") + 
  theme_minimal()
```

Since our model is getting better at predicting the propensity scores, all of our ATO weights will go to zero since more observations are in the groups we predict them to be. 

  
```{r}
smds <- coffee_2 %>%
  summarize(
    across(c(age, job, smoke),
           list(unweighted = ~smd(.x, coffee_town)$estimate,
                ato.weighted = ~smd(.x, coffee_town, w = w_ato)$estimate))
  )

plot_df <- smds %>%
  pivot_longer(
    everything(),
    values_to = "SMD",
    names_to = c("Variable", "Method"),
    names_pattern = "(.*)_(.*)"
  ) %>%
  arrange(Method, abs(SMD)) %>%
  mutate(Variable = fct_inorder(Variable))

ggplot(
  data = plot_df,
  aes(x = abs(SMD), y = Variable, 
      group = Method, color = Method)
) +  
  geom_line(orientation = "y") +
  geom_point() + 
  geom_vline(xintercept = 0.1, 
             color = "black", size = 0.1) +
  theme_minimal()
```

We do not see a change in the Love plot, but that is to be expected since our smd's aren't changing.


```{r}
ecdf_1 <- coffee_2 %>%
  filter(coffee_town == 1) %>%
  arrange(age) %>%
  mutate(cum_pct = cumsum(w_ato) / sum(w_ato))

ecdf_0 <- coffee_2 %>%
  filter(coffee_town == 0) %>%
  arrange(age) %>%
  mutate(cum_pct = cumsum(w_ato) / sum(w_ato))

ggplot(ecdf_1, aes(x = age, y = cum_pct)) +
  geom_line(color = "deeppink2") +
  geom_line(data = ecdf_0, 
            aes(x = age, y = cum_pct), 
            color = "plum4") + 
  xlab("Age") + 
  ylab("Proportion <= x") + 
  ggtitle("Age after ATO Weighting") +
  annotate(geom = "label",
         x = 23,
         y = .64,
         label  = "Lives in Coffee Town",
         color = "deeppink2") + 
  annotate(geom = "label",
           x = 29, 
           y = .5,
           label = "Doesn't Live in Coffee Town",
           color = "plum4") + 
  theme_minimal()
```

This eCDF looks much better than the last ones! We see that the distributions are now very closely matched after iterating our propensity score model. We are satisfied with these plots.

## Exercise 10

```{r}
coffee_2 %>% 
  summarise((sum(coffee_town*cups_of_coffee*w_ato)/sum(coffee_town*w_ato))-(sum((1-coffee_town)*cups_of_coffee*w_ato)/sum((1-coffee_town)*w_ato)))
```
Since the average causal effect is practically zero, this means that among the overlap population, those who live in coffee town do not consume any more cups of coffee than those who do not. 

## Bonus

In order to get an idea of the uncertainty of the average causal effect, we could create a bootstrapped confidence interval. A regular confidence interval won't work since there will be issues with the standard errors being either too large or too small. 
