---
title: "K-Means Clustering"
author: "Ellie Grace Moore"
date: '2022-06-08'
output: 
  html_document: 
    highlight: monochrome
    theme: cosmo
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center')
library(flexclust)
```

```{r}
religion <- read.csv(file="rel.csv")
rownames(religion)=religion[,2]
religion <- religion[,3:19]
head(religion)
```

  For simplicity, we will select only the major religions. With this being said, we will filter out the original `religion` dataset and keep Christian (non-Catholic), Catholic, Jewish, Muslim, and No Religion. 

``` {r}
data <- religion[,c(1,2,3,4,5,7,16)]
head(data)
```

  Now we will determine how many clusters to use for our k-means clustering. 

``` {r}
stepFlexclust(data, k = 2:12, nrep = 20)
plot(stepFlexclust(data, k = 2:12, nrep = 20), type = "l")
```

  Although there is a level of ambiguity in the `stepFlexclust()` function, but according to the graph it appears that the plot appears to begin to "level out" after 5 clusters. Therefore we will select $k=5$. Now we will find the specific cluster assignments that result in the lowest within cluster distance. Below the cluster assingments (for clusters 1-5) for each country is shown below.

```{r}
clustrelbest = stepFlexclust(data,k=5,nrep=1000) 
summary(clustrelbest) 

partition=clusters(clustrelbest)
partition
```