---
title: "Missing Data"
author: "Ellie Grace Moore"
date: "4/14/2022"
output: 
  html_document:
    theme: cosmo
    highlight: kate
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', warning = FALSE)
library(faraway)
library(mice)
library(ggplot2)
library(visdat)
```

Before starting an analysis, one must familiarize themselves with the contents of the dataset. This specifically applies to missing data. Luckily, the `visdat` package allows us to create a plot that outlays the missing data in a visually-friendly way. For this example, we are working with the `mammalsleep` dataset. The first few rows are shown below.

```{r, include=FALSE}
head(mammalsleep)
```

```{r}
vis_dat(mammalsleep, palette = "cb_safe") +
  ggtitle("Visualizing Missing Data for The Mammalsleep Dataset")
```

From the plot above, we are able to tell that nondream and dream have the most missing values, followed by lifespan, sleep, and gestation. A method to "replace" this data is to use predictive mean matching. The code below conducts this process, and we see a new plot to ensure that there are no missing values.

```{r}
mice <- mice(mammalsleep, m=1, method = "pmm", printFlag = FALSE)
complete_data <- complete(mice)
vis_dat(complete_data, palette = "cb_safe") +
  ggtitle("Visualizing Missing Data after Predictive Mean Matching", sub = "There are none!") 
```

