---
title: "Visualizing Distributions"
author: "Ellie Grace Moore"
date: "2/16/2022"
output: 
  html_document:
    theme: cosmo
    highlight: kate
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', messages = FALSE, warning = FALSE)
library(UsingR)
library(car)
library(aplpack)
library(tidyverse)
biomarker <- read_csv("Data/biomarker.csv")
X <- biomarker
```

For this first example, we will work with the `biomarker` dataset. This involves several variables that are related to the health of elderly patients. The variables are the following:

* **PPFEV1**: the forced expiritory volume of air the patient can exhale in one second
* **DSS**: patient scores on mental exercise
* **sixmpace**: speed a person can walk six meters
* **grip**: how tighly a patient can grip an object

The first few rows of the dataframe are shown below. We also include a scatterplot matrix to get a preliminary idea of the variables relation to each other.

```{r, echo=FALSE}
head(X)
S <- cov(X)

scatterplotMatrix(X)
```

None of the variables appear to be strongly related. However, of the ones present we see that grip and sixmpace have the highest positive correlation value of approximately 0.2 and grip and PPFEV1 have the "highest" negative correlation with a value of 0.14.


According to the Q-Q Plots and histograms below, it appears that **DSS** and **grip** least represent a normal distribution. However between these two grip is arguably "less normal" being rather right skewed. 

```{r}
par(mfrow = c(2,2))
qqnorm(biomarker$PPFEV1, main = "Q-Q Plot for PPFEV1")
qqline(biomarker$PPFEV1)
qqnorm(biomarker$DSS, main = "Q-Q Plot for DSS")
qqline(biomarker$DSS)
qqnorm(biomarker$sixmpace, main = "Q-Q Plot for Sixmpace")
qqline(biomarker$sixmpace)
qqnorm(biomarker$grip, main = "Q-Q Plot for Grip")
qqline(biomarker$grip)
```

```{r}
par(mfrow = c(2,2))
hist(biomarker$PPFEV1, main = " Histogram for PPFEV1")
hist(biomarker$DSS, main = "Histogram for DSS")
hist(biomarker$sixmpace, main = "Histogram for Sixmpace")
hist(biomarker$grip, main = "Histogram for Grip")
```

We can now create a kernel density estimate for DSS.

```{r}
hist(biomarker[,2], main="DSS", xlab="DSS", freq=FALSE)
h=seq(0,80,by=1)
v=dnorm(h,mean(biomarker$DSS),sqrt(S[2,2]))
lines(h,v, col="red", lwd=3)
```

In my opinion, since the discrete data appears to follow a normal distribution, I find it okay to use a continuous normal for the `DSS` variable. This way we are able to learn more about the variable itself.

### 4. (a)

This first plot is more "raw" code that allows me to play around with different bandwidths. Then the next chunk of code outputs the optimal bandwidth according to R (and thus giving us a much more viewer-friendly kernel density estimate).

```{r, include = FALSE}
colleges <- read_csv("Data/colleges.csv")
xgrid <- seq(5000,55000,by=1000)
gauss = function(x) 1/sqrt(2*pi) * exp(-(x^2)/2)
```


```{r kde, fig.width=8, fig.height=4, fig.align='center'}
par(mfrow = c(1,2))

x=colleges$TuitionState
n=length(x)
h=1000
bumps=sapply(x, function(a) gauss((xgrid-a)/h)/(n*h))

plot(xgrid, rowSums(bumps), type="l", lwd=2, col="red", main = c("Bandwidth h=",h))
rug(x, lwd=6)
out=apply(bumps, 2, function(b) lines(xgrid, b))
plot(density(colleges$TuitionState), main = "Kernel Density Estimate of \n State Tuition", xlab = "Tuition Cost")

```

To further visulize the data, we will construct bagplots of the graduation rate as well as the fall acceptance rate.

```{r, fig.align='center', fig.width=8, fig.height=4}
par(mfrow=c(1,2))
bagplot(colleges$`4YrGradRate`, xlab = "4YrGradRate", main = "Bagplot of Grad Rate")
bagplot(colleges$Fall14Accep, xlab = "Fall14Accep", main = "Bagplot of Fall Acceptance")
```

Lastly, we can visulize the distribution even *further* by constructing a 3D scatterplot. There is no perfect angle for viewing a 3D Scatterplot, but after looking at different angles, this angle (70 degrees) was determined the most clear.

```{r}
library(scatterplot3d)
with(colleges, scatterplot3d(PercentAid, Fall14Accep, Endowment, type="h", angle=70))

```

